---
title: "West Nile Virus Incidence in Pennsylvania Mosquitos"
author: "Kristina Glass"
output: html_document
date: "2023-12-12"
fig_caption: yes
---

# West Nile Virus Incidence in Pennsylvania Mosquitoes

## Introduction:

In 2018, Pennsylvania experienced one of it's worst West Nile Virus (WNV) outbreaks in 15 years, resulting in around 72 cases reported in humans and three deaths from the virus [1](https://www.publicopiniononline.com/story/news/2018/10/09/pennsylvanias-worst-outbreak-west-nile-virus-15-years/1580572002/). According to the Pennsylvania Department of Environmental Protection (DEP), West Nile Virus first entered the state in 2000 and since then, multiple state departments(Department of Environmental Protection, Department of Agriculture, and Department of Health) have embarked on a joint effort to manage the mosquito population to help reduce the risk of the virus and educate the public on how to reduce their risk of infection thought Integrated Pest Management [2](https://www.dep.pa.gov/Business/ProgramIntegration/Vector-Management/Mosquitoes/Pages/default.aspx).

## What is West Nile Virus?:

West Nile Virus is a vector born disease that cycles itself through birds (hosts) and mosquitoes (vectors), where a bird infected by WNV from a mosquito can in turn infect an uninfected mosquito who then bites that infected bird[2](https://www.dep.pa.gov/Business/ProgramIntegration/Vector-Management/Mosquitoes/Pages/default.aspx). From there, WNV transmits itself to humans and other mammals. In Pennsylvania, the primary concern for WNV transmission in non-human animals are horses and wild game, particularly birds.

## Area of Interest:

Philadelphia has ranked among the top cities for both mosquito rates and WNV cases in humans [1](<https://www.publicopiniononline.com/story/news/2018/10/09/pennsylvanias-worst-outbreak-west-nile-virus-15-years/1580572002/>[)[3](https://www.inquirer.com/science/climate/mosquitos-climate-change-climate-central-global-warming-philadelphia-allentown-atlantic-city-zika-20200804.html](https://www.publicopiniononline.com/story/news/2018/10/09/pennsylvanias-worst-outbreak-west-nile-virus-15-years/1580572002/)[3](https://www.inquirer.com/science/climate/mosquitos-climate-change-climate-central-global-warming-philadelphia-allentown-atlantic-city-zika-20200804.html)). Mosquitoes season ranges from around March to late October when temperatures are warmer, typically seeing spikes in July. But due to climate change, mosquito season is lasting longer, increasing the potential for exposure to WNV. While an increase in mosquitoes does not inherently mean an increase in disease cases, investigating the link could be worthwhile.

## Data:

## Research Question:

What is the impact of temperature and relative humidity on West Nile Virus test results in Philadelphia mosquitoes?

# Analysis

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#load in necessary libraries

library(tidyverse)
library(readr)
library(gt)
library(tufte)
library(feasts)
library(readxl)
library(writexl)
library(janitor)
library(tsibble)
library(sf)
library(raster)
library(gstat)
library(automap)
library(patchwork)
library(viridis)
library(knitr)

options(scipen = 999) # disable scientific notation
```

```{r}
#load in pa mosquito data
data_2018 <- read_excel("data/pa_mosquito_data_2018.xlsx")
#load in philadelphia weather data
philly_weather <- read_excel("data/faa_daily-KPHL_20180516-20181024.xlsx")

#change names of weather data columns
philly_weather <- philly_weather %>% 
        rename("date" = "Date (EST)",
               "avg_temp" = "Average Temp (F)",
               "avg_rel" = "Average Relative Humidity (%)")

#convert columns to lower snake case and drop na values
data_2018<-janitor::clean_names(data_2018, "snake") %>% drop_na()
#subset data to counties of interest
philadelphia <- subset(data_2018, county == 'Philadelphia') 

#rename philadelphia "collected" column to "date"
philadelphia <- philadelphia %>% 
  rename("date" = "collected")
#merge weather data with city mosquito data
philly_merge <- merge(philadelphia,philly_weather,by="date")

#convert results "POS" and "NEG" to factors
philly_merge$result <- as.factor(philly_merge$result)
#confirm change
class(philly_merge$result)
levels(philly_merge$result)
```

## Hypothesis Testing

Here, we conduct a hypothesis test to evaluate the relationships between Average Temperature and Relative Humidity on WNV test results in mosquitoes. Where:

$$H_{0}:$$

***There is no difference in WNV test results based on temperature/relative humidity***

$$H_{A}:$$ ***There is a difference in WNV test results based on temperature/relative humidity***

### Average Temperature

```{r}
#Alternative Hypothesis: There is a difference in wnv result based on average temperature
#Null Hypothesis: There is no difference in wnv result based on average temperature

# compute point estimate of result mean

mu_negative = philly_merge %>% 
  filter(result == "NEG") %>% 
  summarize(mean(avg_temp))

mu_positive = philly_merge %>% 
  filter(result == "POS") %>% 
  summarize(mean(avg_temp))

point_est = as.numeric(mu_negative - mu_positive)

#compute the standard error

n1 = philly_merge %>% filter(result == "NEG") %>% count()
n2 = philly_merge %>% filter(result == "POS") %>% count()
s1 = philly_merge %>% filter(result == "NEG") %>% summarize(sd(avg_temp, na.rm = TRUE))
s2 = philly_merge %>% filter(result == "POS") %>% summarize(sd(avg_temp, na.rm = TRUE))
SE = as.numeric(sqrt(s1^2/n1 + s2^2/n2))

#compute test statistic
zscore = (point_est - 0)/SE
print(zscore)

#print(paste0("The observed difference in average temperature is [" zscore "] standard deviations bellow the mean of 0"))

#compute the critical value at a 95% confidence interval
crit_val = qnorm(0.025, lower.tail=FALSE)
#find the upper and lower confidence intervals
ci_lower = round(point_est - crit_val*SE, 2)
ci_upper = round(point_est + crit_val*SE, 2)

print(paste0("there is a 95% probability that [", ci_lower, " ,", ci_upper, "] contains the difference in means between WNV results"))
```

```{r}
#conduct canned t-test to confirm manual test worked
t.test(avg_temp ~ result, philly_merge)
#find summary of linar model of (x) on (y)
summary(lm(avg_temp ~ result, philly_merge))
```

### Average Relative Humidity

```{r}
#Alternative Hypothesis: There is a difference in wnv result based on average temperature
#Null Hypothesis: There is no difference in wnv result based on average temperature

# compute point estimate of result mean

mu_negative2 = philly_merge %>% 
  filter(result == "NEG") %>% 
  summarize(mean(avg_rel))

mu_positive2 = philly_merge %>% 
  filter(result == "POS") %>% 
  summarize(mean(avg_rel))

point_est2 = as.numeric(mu_negative2 - mu_positive2)

#compute the standard error

n3 = philly_merge %>% filter(result == "NEG") %>% count()
n4 = philly_merge %>% filter(result == "POS") %>% count()
s3 = philly_merge %>% filter(result == "NEG") %>% summarize(sd(avg_rel, na.rm = TRUE))
s4 = philly_merge %>% filter(result == "POS") %>% summarize(sd(avg_rel, na.rm = TRUE))
SE2 = as.numeric(sqrt(s1^2/n1 + s2^2/n2))

#compute test statistic
zscore2 = (point_est2 - 0)/SE2
print(zscore2)

#print(paste0("The observed difference in average temperature is [" zscore "] standard deviations bellow the mean of 0"))

#compute the critical value at a 95% confidence interval
crit_val2 = qnorm(0.025, lower.tail=FALSE)
#find the upper and lower confidence intervals
ci_lower2 = round(point_est2 - crit_val2*SE2, 2)
ci_upper2 = round(point_est2 + crit_val2*SE2, 2)

print(paste0("there is a 95% probability that [", ci_lower2, " ,", ci_upper2, "] contains the difference in means between WNV results"))
```

```{r}
#conduct canned t_test to confirm manual results are the same
print(t.test(avg_rel ~ result, philly_merge))
#check summary for linear model
print(summary(lm(avg_rel ~ result, philly_merge)))
```

## Plotting Linear Model

### Average Temperature

```{r, fig.cap = "Positive and Negative test results for West Nile Virus Cases in Philadelphia, PA in 2018. Data was collected from May 16th to October 24th"}
#plot density distribtion of WNV Result on Avg Temp
ggplot(philly_merge, aes(x = result, y = avg_temp)) +
  geom_violin(fill = "steelblue") +
  geom_boxplot(width = .15, 
               fill = "darksalmon",
               outlier.color = "darksalmon",
               outlier.size = 1,
               outlier.alpha = 0.5) + 
  labs(title = "West Nile Virus Test Result Distribution by Average Temp",
       y = "Average Temp (F)",
       x = "Test Results") +
  theme_minimal()

```

### Average Relative Humidity

```{r}
#plot WNV on relative humidity
ggplot(philly_merge, aes(x = result, y = avg_rel)) +
  geom_violin(fill = "steelblue") +
  geom_boxplot(width = .15, 
               fill = "darksalmon") + 
  labs(title = "West Nile Virus Test Result Distribution by Average Relative Humidity",
       y = "Average Relative Humidity (%)",
       x = "Test Results")
```

## Time Series Analysis

```{r}
#aggregate count of mosquitos by date to combine overlapping dates
philadelphia_sorted<-aggregate(count ~ date, data = philadelphia, FUN = "sum")

#plot basic time series analysis by day per month of collection
philadelphia_sorted %>% 
  ggplot(aes(x=date, y = count)) +
  geom_line()+ #daily plot is too hight res to see anyting
  #mutate(
    #yr_mo = yearmonth(date),
    #month = month(date)
  #) %>% 
  #ggplot(aes(yr_mo, count)) +
  stat_summary(geom = 'line', fun = "mean", alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE, color = 'midnightblue')+
  labs(x='Month of Collection', y="Mosquito Counts",
       title = "Distribution of Mosquito Counts Over Month by Date of Collection")
```

```{r}
#plot basic time series by month of collection
philadelphia_sorted %>% 
  #ggplot(aes(x=date, y = count)) +
  #geom_line()+ #daily plot is too hight res to see anyting
  mutate(
    yr_mo = yearmonth(date),
    month = month(date)
  ) %>% 
  ggplot(aes(yr_mo, count)) +
  stat_summary(geom = 'line', fun = "mean", alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE, color = 'midnightblue')+
  labs(x='Month of Collection', y="Mosquito Counts",
       title ="Distribution of Mosquito counts by Time Over Month")
```

```{r}
#turn dataframe into time series object
philly_ts = philadelphia_sorted %>% 
  mutate(yr_mo = yearmonth(date)) %>% 
  group_by(yr_mo) %>% 
  summarise(count_mean = mean(count, na.rm = TRUE)) %>% 
  tsibble(index = yr_mo) 

#check head of object
philly_ts %>% head(10)            
            
```

```{r}
#decompose timeseries object
dcmp <- philly_ts %>% 
  model(STL(count_mean))

#obtain the components from dcmp
components(dcmp) %>% 
  autoplot()
```

```{r}
#plot components of decomposition
components(dcmp) %>% 
  as_tsibble() %>%  #make it a TS Table
  autoplot(count_mean, color = 'grey') + #original data
  # trend + noise
  geom_line(aes(y = trend), color = 'blue') +
  labs(x = 'Year', y = 'Monthly Mean Temp',
       title = "Time Series Decomposition of Mosquito Counts per Month")
```

## Spatial Interpolation and Kriging

```{r}
#plot 
data_2018 %>% 
  ggplot(mapping = aes(x = longitude, y = latitude, color = count)) +
  geom_point(size = 1, alpha = 0.5) + 
  scale_color_viridis(option = "B") + 
  theme_classic() +
labs(title = "Spatial Distribution of Mosquito Counts")
```

```{r}
pa_krig<-data_2018 %>%
   arrange(latitude, longitude) %>%
   distinct(latitude, longitude, .keep_all = TRUE)
```

```{r}
pa_sf <- st_as_sf(pa_krig, coords = c("longitude", "latitude"), crs = 6565) %>% 
  cbind(st_coordinates(.))

head(pa_sf)
```

```{r}
v_mod_full <- automap:::autofitVariogram(count~1, pa_sf)

v_mod <- v_mod_full$var_model
head(v_mod)
```

```{r}
plot(v_mod_full)
```

```{r}
grd_sf <- st_make_grid(pa_sf, cellsize = .5)
plot(grd_sf)

```

```{r}
ok <- krige(
  count~1,
  pa_sf,
  grd_sf,
  model=v_mod)
```

```{r}
dim(ok)
head(ok$var1.pred)
head(ok$var1.var)
```

```{r}
p_raw <- ggplot(data = data_2018,
                mapping = aes(x=longitude, y=latitude, color=count)) +
  geom_point(size = 1) + scale_color_viridis(option = 'B') +
  ggtitle(label = "Observation wells") +
  theme_void()

p_kriging <-  ggplot(ok, aes(x = x, y = y, fill = var1.pred)) +
    geom_raster() +
    ggtitle(label = "Ordinary Kriging") +
    scale_fill_viridis(option = "B") +
    theme_void() +
    theme(
      plot.title = element_text(hjust = 0.5)
    ) + guides(fill=guide_legend(title="Z", reverse=TRUE))

p_raw + p_kriging
```
